---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Every other way of doing chess in R starts from a position you already have.
tanmai starts one step earlier: it will read the position off a picture.

```{r}
library(tanmai)
```

## Four ways in, two shapes out

There are four ways to get a position into R, and they all produce one of two
data frames. Nothing downstream asks which one you used.

| | gives you |
| --- | --- |
| `read_fen()` | a position |
| `read_board()` | a position, plus how confident it is |
| `read_pgn()` | a game |
| `read_video()` | a game |

A **position** is one row. A **game** is one row per ply:

```{r}
game <- read_pgn("1. e4 e5 2. Nf3 Nc6 3. Bb5 a6")
names(game)
#> [1] "ply" "move_no" "side" "san" "uci" "fen_before" "fen_after"
#> [8] "cp" "cp_loss" "best_uci" "best_san" "class" "human_p"
```

They are plain data frames. `[`, dplyr and ggplot2 all work on them without
knowing anything about chess, and that is the point.

## Reading a board from a picture

```{r}
pos <- read_board("screenshot.png")
pos$fen
```

This is the only input that can be *wrong*, so it is the only one that reports
how sure it is:

```{r}
pos$confident
#> TRUE
```

When `confident` is `FALSE`, the board did not match any installed piece set
well - usually a Chess.com or custom theme. The FEN still comes back, because a
guess you can correct beats no answer, but it is flagged rather than presented
as fact.

Out of the box the bundled set is recognised. `setup_piece_sets()` fetches the
Lichess sets, after which most screenshots from most sites are read with
nothing to configure.

### If you already have the FEN

Then recognition is a step to skip:

```{r}
read_fen("r1bqk1nr/pppp1ppp/2n5/2b1p3/2B1P3/5N2/PPPP1PPP/RNBQK2R w KQkq -")
```

Four fields is fine. That is what chess sites actually copy to the clipboard,
and the rest are filled in - though castling is deliberately *not* invented: a
FEN that omits the field is not claiming those rights.

## Reading a game from a recording

```{r}
game <- read_video("my-game.mp4")
```

A screen recording, an animated GIF, or a folder of screenshots all work. GIFs
and folders need nothing installed; video formats need ffmpeg.

A recording is full of frames that must not be believed - a piece halfway
through a slide, the board mid-redraw, a piece picked up and put back down. A
frame joins the game only when **exactly one legal sequence of moves explains
it**, so those are discarded rather than turned into invented moves. What
happened to each frame is recorded rather than hidden:

```{r}
attr(game, "ledger")
#>       outcome frames
#> 1     skipped     12
#> 2    accepted     10
#> 3 unexplained      1
```

## Judging a game

Everything above works with no chess engine at all. Judging a game needs one:

```{r}
has_engine()
```

If that is `TRUE`:

```{r}
game <- evaluate(read_pgn(my_pgn))
game[, c("ply", "san", "cp", "cp_loss", "class")]
```

`cp` is always from **White's** point of view, so the sign means the same thing
down the whole column. `cp_loss` is from the **mover's**, because "you lost
three pawns" only reads correctly that way for both colours.

```{r}
accuracy(game)
#>   side moves  acpl accuracy inaccuracies mistakes blunders
#> 1    w    23  58.6     84.3            3        2        2
#> 2    b    22 513.5     77.5            3        2        3

turning_points(game)
```

The accuracy figure uses Lichess's published formula, so it is comparable with
the number they show. It converts each evaluation to an expected win percentage
first, which is the point: half a pawn matters enormously in a level position
and not at all when you are already winning by a rook.

## Drawing

```{r}
plot(read_fen(fen)) # the board
plot(game) # the evaluation across the game
```

## What it cannot do

- **Whose turn it is** cannot be read from a still image, so `read_board()` has
  to be told.
- **En passant** is not inferred, and castling rights only when king and rook
  are on their home squares.
- A few Lichess piece sets are genuinely hard to tell apart because their own
  pieces are nearly identical; `mono`, `letter` and `disguised` are excluded.
